---
title: "MP01: Netflix"
---

```{r}
#| echo: false
library(readr)
library(dplyr)

# Data Import
if(!dir.exists(file.path("data", "mp01"))){
  dir.create(file.path("data", "mp01"), showWarnings=FALSE, recursive=TRUE)
}

# File paths to tsv
GLOBAL_TOP_10_FILENAME <- file.path("data", "mp01", "global_top10_alltime.tsv")
COUNTRY_TOP_10_FILENAME <- file.path("data", "mp01", "country_top10_alltime.tsv")
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
# Downloading global data
if(!file.exists(GLOBAL_TOP_10_FILENAME)){
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-global.tsv",
                destfile=GLOBAL_TOP_10_FILENAME)
}

# Download country data if missing
if(!file.exists(COUNTRY_TOP_10_FILENAME)){
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-countries.tsv",
                destfile=COUNTRY_TOP_10_FILENAME)
}

# Read the data into R, treating "N/A" as proper NA
GLOBAL_TOP_10 <- read_tsv(GLOBAL_TOP_10_FILENAME, na = "N/A")
COUNTRY_TOP_10 <- read_tsv(COUNTRY_TOP_10_FILENAME, na = "N/A")

# Quick checks
glimpse(GLOBAL_TOP_10)
glimpse(COUNTRY_TOP_10)
