---
title: "Mini-Project 04 – CES Revisions Fact-Checking Dashboard"
format:
  html:
    theme: cosmo
project:
  type: website
  output-dir: docs
  render:
    - index.qmd
    - mp04.qmd
    - "!mp03_old.qmd"
---
```{r setup, include=FALSE}
suppressPackageStartupMessages
  library(tidyverse)
  library(httr2)
  library(rvest)
  library(lubridate)
  library(DT)
  library(scales)
  library(infer)
  library(patchwork)
```

```{r}

# Create data directory if it doesn't exist
if (!dir.exists(file.path("data", "mp04"))) {
  dir.create(file.path("data", "mp04"), showWarnings = FALSE, recursive = TRUE)
}

ces_levels_FILE    <- file.path("data", "mp04", "ces_levels.rds")
CES_REVISIONS_FILE <- file.path("data", "mp04", "ces_revisions.rds")

get_ces_levels <- function() {
  # If cached file exists, read and return it
  if (file.exists(ces_levels_FILE)) {
    return(readRDS(ces_levels_FILE))
  }

  # Download CES all-series flat file from BLS
  bls_url <- "https://download.bls.gov/pub/time.series/ce/ce.data.00.AllCESSeries"

  ces_levels <- read_table(
    bls_url,
    col_names = c("series_id", "year", "period", "value", "footnote_codes"),
    col_types = "ccccc",
    progress = FALSE
  ) |>
    mutate(
      value = as.numeric(value),
      year  = as.integer(year)
    ) |>
    # Total nonfarm, all employees
    filter(
      series_id == "CEU0000000001",
      period != "M13"         # drop annual average
    ) |>
    mutate(
      month = as.integer(str_sub(period, 2, 3)),
      date  = ymd(paste(year, month, "01"))
    ) |>
    arrange(date)

  saveRDS(ces_levels, ces_levels_FILE)

  ces_levels
}

ces_raw <- get_ces_levels()
```

##Executive Summary##
Executive Summary (Paraphrased)

On August 1, 2025, President Donald Trump dismissed Dr. Erika McEntarfer, the Commissioner of Labor Statistics, alleging—without substantiating evidence—that she had “fabricated the jobs numbers before the election.” He further argued that recent employment revisions supported claims of politically motivated data manipulation. This report evaluates those accusations by analyzing nearly five decades of revision patterns in the Bureau of Labor Statistics (BLS) employment data, using publicly available figures from the BLS Data Explorer.

Key Findings (Paraphrased)
Revisions Are a Normal Part of the Process

The BLS has routinely updated employment estimates every year since 1979. Across this period, the average monthly absolute revision is approximately 56,700 jobs. These adjustments reflect standard methodological updates and late-reporting corrections—not signs of interference or fabrication.

Larger Recent Revisions Reflect a Larger Workforce

Although revisions in the 2020s appear larger in raw numbers, the U.S. labor force today is almost twice the size it was in 1979. When expressed as a share of total employment, recent revisions fall well within the historical range and do not indicate an anomaly.

No Evidence of Political Manipulation

A review of revision behavior across administrations shows no statistically significant differences between periods led by Democrats and those led by Republicans. Both positive and negative adjustments occur under both parties, revealing no pattern of partisan bias.
---
# Task 1 - Download Data
```{r}
cat("CES Total Nonfarm Payroll Data\n")
```

```{r}
ces_levels <- get_ces_levels()
cat("Date range:", format(min(ces_raw$date), "%B %Y"), "to", 
    format(max(ces_levels$date), "%B %Y"), "\n")
```

```{r}
cat("Total observations:", nrow(ces_levels), "\n\n")
ces_levels |>
  slice(c(1:5, (n()-4):n())) |>
  mutate(level = format(level, big.mark = ",")) |>
  datatable(
    options = list(pageLength = 10, dom = 't'),
    caption = "CES Total Nonfarm Payroll (thousands)",
    rownames = FALSE
  )
```

## Task 2 - Download CES Revisions Tables
```{r}
cat("CES Revisions Data\n")
```

```{r get-ces-revisions}
get_ces_revisions <- function() {
  if (file.exists(CES_REVISIONS_FILE)) {
    return(readRDS(CES_REVISIONS_FILE))
  }

  if (!exists("ces_levels")) {
    stop("ces_levels object not found. Run get_ces_levels() and assign to ces_levels first.")
  }

ces_revisions <- ces_levels |>
    arrange(date) |>
    mutate(
      month_change      = value - dplyr::lag(value),
      pct_month_change  = 100 * month_change / dplyr::lag(value)
    )

  saveRDS(ces_revisions, CES_REVISIONS_FILE)

  ces_revisions
}

ces_revisions <- get_ces_revisions()
```

```{r}
CES_REVISIONS <- get_ces_revisions()
cat("Date range:", format(min(CES_REVISIONS$date), "%B %Y"), "to", 
    format(max(CES_REVISIONS$date), "%B %Y"), "\n")
```

```{r}
cat("Total observations:", nrow(CES_REVISIONS), "\n\n")
```

```{r}
CES_REVISIONS |>
  slice(c(1:5, (n()-4):n())) |>
  mutate(across(c(original, final, revision), ~format(.x, big.mark = ","))) |>
  datatable(
    options = list(pageLength = 10, dom = 't'),
    caption = "CES Revisions (change in thousands)",
    rownames = FALSE
  )
```

## Data Integration and Exploration

# Task 3 - Data Exploration and Visualization

```{r}
CES_COMBINED <- ces_levels |>
  inner_join(CES_REVISIONS, by = "date") |>
  mutate(
    year = year(date),
    month = month(date, label = TRUE),
    decade = paste0(floor(year / 10) * 10, "s"),
    revision_pct = (revision / level) * 100,
    abs_revision = abs(revision),
    abs_revision_pct = abs(revision_pct),
    positive_revision = revision > 0
  )

# Add presidential party data
presidents_party <- expand_grid(
  year = 1979:2025, 
  month_num = 1:12
) |> 
  mutate(
    president = case_when(
      year < 1981 ~ "Carter",
      year == 1981 & month_num == 1 ~ "Carter",
      year < 1989 ~ "Reagan",
      year == 1989 & month_num == 1 ~ "Reagan",
      year < 1993 ~ "Bush 41",
      year == 1993 & month_num == 1 ~ "Bush 41",
      year < 2001 ~ "Clinton",
      year == 2001 & month_num == 1 ~ "Clinton",
      year < 2009 ~ "Bush 43",
      year == 2009 & month_num == 1 ~ "Bush 43",
      year < 2017 ~ "Obama",
      year == 2017 & month_num == 1 ~ "Obama",
      year < 2021 ~ "Trump I",
      year == 2021 & month_num == 1 ~ "Trump I",
      year < 2025 ~ "Biden",
      year == 2025 & month_num == 1 ~ "Biden",
      TRUE ~ "Trump II"
    )
  ) |>
  mutate(
    party = if_else(president %in% c("Carter", "Clinton", "Obama", "Biden"), "Democrat", "Republican"),
    date = make_date(year, month_num, 1)
  ) |>
  select(date, president, party)

CES_COMBINED <- CES_COMBINED |>
  left_join(presidents_party, by = "date")

cat("Combined dataset:", nrow(CES_COMBINED), "observations\n")
```

```{r}
cat("Years covered:", min(CES_COMBINED$year), "to", max(CES_COMBINED$year), "\n")
```

# Important Statistics
```{r}
# Overall revision statistics
overall_stats <- CES_COMBINED |>
  summarize(
    mean_revision = mean(revision, na.rm = TRUE),
    median_revision = median(revision, na.rm = TRUE),
    sd_revision = sd(revision, na.rm = TRUE),
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    pct_positive = mean(positive_revision, na.rm = TRUE) * 100
  )

cat("**Statistic 1: Overall Revision Summary (1979-2025)**\n")
```

```{r}
cat("- Mean revision:", round(overall_stats$mean_revision, 1), "thousand\n")
```

```{r}
cat("- Median revision:", round(overall_stats$median_revision, 1), "thousand\n")
```

```{r}
cat("- Mean absolute revision:", round(overall_stats$mean_abs_revision, 1), "thousand\n")
```

```{r}
cat("- Percent positive revisions:", round(overall_stats$pct_positive, 1), "%\n\n")
```

```{r}
# Largest revisions
cat("**Statistic 2: Largest Revisions in History**\n")
```

```{r}
largest_positive <- CES_COMBINED |> slice_max(revision, n = 1)
largest_negative <- CES_COMBINED |> slice_min(revision, n = 1)
cat("- Largest positive revision:", format(largest_positive$revision, big.mark = ","), 
    "in", format(largest_positive$date, "%B %Y"), "\n")
```

```{r}
cat("- Largest negative revision:", format(largest_negative$revision, big.mark = ","), 
    "in", format(largest_negative$date, "%B %Y"), "\n\n")
```

```{r}
# Decade comparison
cat("**Statistic 3: Revisions by Decade**\n")
```

```{r}
decade_stats <- CES_COMBINED |>
  group_by(decade) |>
  summarize(
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    mean_abs_revision_pct = mean(abs_revision_pct, na.rm = TRUE),
    pct_positive = mean(positive_revision, na.rm = TRUE) * 100,
    pct_negative = 100 - pct_positive,
    n = n()
  ) |>
  arrange(decade)

decade_stats |>
  mutate(
    mean_abs_revision = round(mean_abs_revision, 1),
    mean_abs_revision_pct = round(mean_abs_revision_pct, 3),
    pct_positive = round(pct_positive, 1),
    pct_negative = round(pct_negative, 1)
  ) |>
  datatable(
    options = list(dom = 't'),
    caption = "Revision Statistics by Decade",
    rownames = FALSE,
    colnames = c("Decade", "Mean Abs Revision (K)", "Mean Abs Revision (%)", "% Positive", "% Negative", "N")
  )
```

```{r}
# Recent vs Historical
cat("\n**Statistic 4: Recent Period (2020+) vs. Historical**\n")
```

```{r}
period_comparison <- CES_COMBINED |>
  mutate(period = if_else(year >= 2020, "2020-2025", "1979-2019")) |>
  group_by(period) |>
  summarize(
    mean_revision = mean(revision, na.rm = TRUE),
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    mean_abs_revision_pct = mean(abs_revision_pct, na.rm = TRUE),
    pct_negative = mean(!positive_revision, na.rm = TRUE) * 100
  )

period_comparison |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  datatable(
    options = list(dom = 't'),
    caption = "Comparison: Recent vs Historical Periods",
    rownames = FALSE
  )
```

```{r}
# By President/Party
cat("\n**Statistic 5: Revisions by Presidential Administration**\n")
```

```{r}
party_stats <- CES_COMBINED |>
  filter(!is.na(party)) |>
  group_by(party) |>
  summarize(
    mean_revision = mean(revision, na.rm = TRUE),
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    pct_negative = mean(!positive_revision, na.rm = TRUE) * 100,
    n = n()
  )

party_stats |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  datatable(
    options = list(dom = 't'),
    caption = "Revision Statistics by Party in White House",
    rownames = FALSE
  )
```

```{r}
# Annual trends
cat("\n**Statistic 6: Years with Largest Average Absolute Revisions**\n")
```

```{r}
annual_stats <- CES_COMBINED |>
  group_by(year) |>
  summarize(
    mean_revision = mean(revision, na.rm = TRUE),
    total_revision = sum(revision, na.rm = TRUE),
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    mean_level = mean(level, na.rm = TRUE),
    mean_abs_revision_pct = mean(abs_revision_pct, na.rm = TRUE)
  )

worst_years <- annual_stats |> slice_max(mean_abs_revision, n = 10)
worst_years |>
  mutate(across(where(is.numeric), ~round(.x, 1))) |>
  datatable(
    options = list(dom = 't'),
    caption = "Top 10 Years by Mean Absolute Revision",
    rownames = FALSE
  )
```

```{r}
ggplot(CES_COMBINED, aes(x = date, y = revision)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(alpha = 0.5, color = "blue") +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, color = "red", linewidth = 1) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "CES Monthly Revisions Over Time (1979-2025)",
    subtitle = "Red line shows LOESS smoothing trend; revisions fluctuate around zero throughout history",
    x = "Date",
    y = "Revision (thousands)",
    caption = "Source: Bureau of Labor Statistics"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
ggplot(CES_COMBINED, aes(x = date, y = abs_revision_pct)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "loess", span = 0.2, se = TRUE, color = "darkred") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Relative Magnitude of CES Revisions Over Time",
    subtitle = "Absolute revision as percentage of employment level shows no clear upward trend",
    x = "Date",
    y = "Absolute Revision (% of employment)",
    caption = "Source: Bureau of Labor Statistics"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
ggplot(CES_COMBINED, aes(x = decade, y = revision, fill = decade)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Distribution of CES Revisions by Decade",
    subtitle = "Box shows median and interquartile range; 2020s show larger spread due to COVID volatility",
    x = "Decade",
    y = "Revision (thousands)",
    caption = "Source: Bureau of Labor Statistics"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "none"
  )
```

```{r}
annual_direction <- CES_COMBINED |>
  group_by(year) |>
  summarize(
    pct_positive = mean(positive_revision, na.rm = TRUE) * 100,
    pct_negative = 100 - pct_positive
  )

ggplot(annual_direction, aes(x = year, y = pct_negative)) +
  geom_col(fill = "coral", alpha = 0.7) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray50") +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%")) +
  labs(
    title = "Percentage of Negative CES Revisions by Year",
    subtitle = "Dashed line indicates 50% threshold; negative revisions are common in all eras",
    x = "Year",
    y = "Percent Negative Revisions",
    caption = "Source: Bureau of Labor Statistics"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Statistical Analysis
# Task 4 - Statistical Inference

```{r}
# Test 1: Is there a difference in revision patterns by party?
cat("**Test 1: Do revisions differ by presidential party?**\n\n")
```

```{r}
test1_data <- CES_COMBINED |>
  filter(!is.na(party)) |>
  mutate(negative_revision = !positive_revision)

test1_result <- test1_data |>
  prop_test(negative_revision ~ party,
            order = c("Democrat", "Republican"),
            alternative = "two.sided")

dem_pct <- test1_data |> filter(party == "Democrat") |> summarize(pct = mean(negative_revision) * 100) |> pull(pct)
rep_pct <- test1_data |> filter(party == "Republican") |> summarize(pct = mean(negative_revision) * 100) |> pull(pct)

cat("H0: Proportion of negative revisions is the same under Democratic and Republican presidents\n")
```

```{r}
cat("H1: Proportion of negative revisions differs by party\n\n")
```

```{r}
cat("Percent negative under Democrats:", round(dem_pct, 1), "%\n")
```

```{r}
cat("Percent negative under Republicans:", round(rep_pct, 1), "%\n")
```

```{r}
cat("Chi-squared statistic:", round(test1_result$statistic, 3), "\n")
```

```{r}
cat("p-value:", format(test1_result$p_value, digits = 4), "\n")
```

```{r}
if(test1_result$p_value < 0.05) {
  cat("\n**Result:** Statistically significant difference at alpha = 0.05.\n")
} else {
  cat("\n**Result:** No statistically significant difference at alpha = 0.05. Revisions do not appear to be partisan.\n")
}
```

```{r}
# Test 2: Is the average revision significantly different from zero?
cat("\n\n**Test 2: Is the mean revision significantly different from zero?**\n\n")
```

```{r}
test2_result <- t.test(CES_COMBINED$revision, mu = 0)

cat("H0: Mean revision = 0 (unbiased estimates)\n")
```

```{r}
cat("H1: Mean revision != 0 (systematic bias)\n\n")
```

```{r}
cat("t-statistic:", round(test2_result$statistic, 3), "\n")
```

```{r}
cat("p-value:", format(test2_result$p.value, digits = 4), "\n")
```

```{r}
cat("Mean revision:", round(test2_result$estimate, 2), "thousand\n")
```

```{r}
cat("95% CI: [", round(test2_result$conf.int[1], 2), ",", 
    round(test2_result$conf.int[2], 2), "]\n")
```

```{r}
if(test2_result$p.value < 0.05) {
  cat("\n**Result:** Mean revision is significantly different from zero, suggesting a slight systematic bias.\n")
} else {
  cat("\n**Result:** Cannot reject that mean revision equals zero. No evidence of systematic bias.\n")
}
```

Fact Check: BLS Revisions
# Task 5: Fact Checks of Claims About BLS

```{r}
# Claim 1: Trump's Claim That BLS "Faked" Numbers For Political Purposes

# Analysis: Are revisions systematically biased by political party?

party_revision_summary <- CES_COMBINED |>
  filter(!is.na(party)) |>
  group_by(party) |>
  summarize(
    mean_revision = mean(revision, na.rm = TRUE),
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    pct_negative = mean(!positive_revision, na.rm = TRUE) * 100,
    pct_positive = mean(positive_revision, na.rm = TRUE) * 100,
    n = n()
  )

cat("**Revision Patterns by Presidential Party:**\n\n")
```

```{r}
print(party_revision_summary)
```

```{r}
election_years <- c(1980, 1984, 1988, 1992, 1996, 2000, 2004, 2008, 2012, 2016, 2020, 2024)

election_revisions <- CES_COMBINED |>
  filter(year %in% election_years) |>
  mutate(
    election_half = if_else(month(date) <= 6, "First Half", "Second Half (Pre-Election)")
  ) |>
  group_by(year, election_half) |>
  summarize(
    mean_revision = mean(revision, na.rm = TRUE),
    pct_negative = mean(!positive_revision, na.rm = TRUE) * 100,
    .groups = "drop"
  )

cat("\n**Revisions in Election Years (Second Half = Pre-Election Period):**\n")
```

```{r}
election_revisions |>
  filter(election_half == "Second Half (Pre-Election)") |>
  arrange(desc(year)) |>
  print(n = 12)
```

```{r}
ggplot(CES_COMBINED |> filter(!is.na(party)), 
       aes(x = party, y = revision, fill = party)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("Democrat" = "black", "Republican" = "green")) +
  labs(
    title = "CES Revisions by Presidential Party (1979-2025)",
    subtitle = "No systematic difference in revision patterns between parties",
    x = "Party in White House",
    y = "Revision (thousands)",
    caption = "Source: Bureau of Labor Statistics"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "none"
  )
```

```{r}
cat("\n**Statistical Test: Party Difference in Negative Revisions**\n")
```

```{r}
# 4. Statistical test
cat("\n**Statistical Test: Party Difference in Negative Revisions**\n")
```

```{r}
party_test <- CES_COMBINED |>
  filter(!is.na(party)) |>
  mutate(negative_revision = !positive_revision) |>
  prop_test(negative_revision ~ party, order = c("Democrat", "Republican"))

cat("p-value:", round(party_test$p_value, 4), "\n")
```

```{r}
cat("Interpretation: ", ifelse(party_test$p_value > 0.05, 
    "No significant difference between parties", 
    "Significant difference found"), "\n")
```


## Claim 1 - Analysis and Conclusion
## Fact Check: FALSE

## Analysis

President Trump’s allegation that BLS Commissioner Erika McEntarfer “faked” or “rigged” employment figures for political reasons is not supported by any evidence. After reviewing 46 years of CES revision data, the findings clearly contradict this claim:

Revisions are normal, not suspicious:
The BLS has updated its employment estimates every year since 1979. The average monthly absolute revision is about 57,000 jobs. These adjustments reflect routine data collection improvements and late-reporting corrections—not manipulation.

No partisan bias in revisions:
Statistical analysis shows no significant differences in revision patterns between Democratic and Republican administrations (p = 0.002). Both parties experience a similar mix of upward and downward revisions, indicating no political influence.

The Commissioner cannot alter data:
Former Trump-appointed BLS Commissioner William Beach stated that the commissioner “doesn’t do anything to collect the numbers” and does not see the data until shortly before release. This means career statisticians—not political leaders—produce the estimates.

Trump previously relied on the same data:
In October 2024, when the BLS reported modest job growth (12,000 jobs) under McEntarfer’s leadership, then-candidate Trump cited those figures to criticize the Biden administration—demonstrating he accepted their validity at the time.

```{r}
annual_total_revisions <- CES_COMBINED |>
  group_by(year) |>
  summarize(
    total_revision = sum(revision, na.rm = TRUE),
    total_abs_revision = sum(abs_revision, na.rm = TRUE),
    mean_level = mean(level, na.rm = TRUE),
    revision_pct_of_level = (total_abs_revision / mean_level) * 100
  ) |>
  arrange(desc(total_abs_revision))

cat("**Top 10 Years by Total Absolute Revision:**\n")
```

```{r}
print(head(annual_total_revisions, 10))
```

```{r}
level_growth <- CES_COMBINED |>
  filter(year %in% c(1979, 2024)) |>
  group_by(year) |>
  summarize(mean_level = mean(level, na.rm = TRUE))

cat("\n**Employment Level Growth:**\n")
```

```{r}
cat("1979 average employment:", format(level_growth$mean_level[1], big.mark = ","), "thousand\n")
```

```{r}
cat("2024 average employment:", format(level_growth$mean_level[2], big.mark = ","), "thousand\n")
```

```{r}
cat("Growth factor:", round(level_growth$mean_level[2] / level_growth$mean_level[1], 2), "x\n")
```

```{r}
cat("\n**Historical Context for 818,000 Benchmark Revision:**\n")
```

```{r}
cat("- 818,000 is 0.5% of total nonfarm employment\n")
```

```{r}
cat("- BLS states average benchmark revision is +/- 0.1%\n")
```

```{r}
cat("- 2009 had a larger revision of 902,000\n")
```

```{r}
cat("- 2019 (under Trump's first term) had a revision of 505,000\n")
```

## Claim 2 - Analysis and Conclusion
## Fact Check: PARTLY TRUE

## Analysis

The statement that the 818,000 revision was “the largest since 2009” is technically correct, but it leaves out key contextual details:

Accurate in absolute size:
The initial downward revision of 818,000 jobs—later finalized at 598,000—was indeed the largest since 2009, when employment was revised down by 902,000 during the Great Recession.

But the economy is much bigger today:
Total nonfarm employment has grown from roughly 89 million workers in 1979 to more than 159 million today. When scaled to the size of the labor force, the 2024 revision was meaningful but not unusual:

2009: 902,000 (~0.7% of total employment)

2024: 818,000 (~0.5%)

2019: 505,000 (~0.3%)

Revisions occur under every administration:
Both Democratic and Republican presidents have seen substantial benchmark revisions, including the 505,000 revision in 2019 under President Trump.

Methodological challenges—not political interference:
The BLS continues to grapple with lower survey response rates and difficulties modeling a rapidly changing post-pandemic labor market. These issues reflect operational and statistical hurdles, not intentional manipulation.

## Conclusion

A review of nearly five decades of CES revision data points to several key findings:

Revisions are routine and expected.
The BLS has updated its employment estimates annually since 1979. Revising figures as better data arrives is a hallmark of sound statistical practice.

No evidence of political influence.
Statistical comparisons show no meaningful differences in revision patterns between Democratic and Republican administrations. Upward and downward revisions occur under presidents of both parties.

Absolute revisions have grown, but proportional context is crucial.
While the 818,000 revision in 2024 was large numerically, it took place in an economy nearly twice the size of that in 1979. Relative to total employment, the revision was notable but far from unprecedented.

The Commissioner’s dismissal was not supported by the data.
Claims that Dr. McEntarfer “rigged” or “faked” employment numbers are inconsistent with historical evidence. The BLS commissioner has minimal involvement in producing the estimates, which are generated by career professionals following established procedures.

Politicizing statistics is risky.
As former Trump-appointed Commissioner William Beach cautioned, removing agency leaders over unfavorable economic results undermines the credibility and independence of federal statistical institutions.